---
title: "bootcamp-r-problem-set-1"
author: "Insert your name here"
output: html_document
---

```{r, setup}
library(tidyverse)
library(knitr)
```

### Problem Set

Each problem below is worth **4 points**.

Use the data files in the `data/` directory to answer the questions.

For this problem set, you are allowed to help each other, but you are
not allowed to post correct answers in slack.

**The problem set is due 12pm on Aug 26.**

### Grading rubric

-   Everything is good: 4 points.
-   Runs, but wrong answer: 3 points
-   Reasonable attempt: 2 points

### Question 1 - Importing dataset

Import the dataset `data_transcript_exp_subset` using the *readr*
package

Hint: The file is located at the following path
`data/data_transcript_exp_subset.csv.gz`

```{r, q1}
```

### Question 2 - Explore the dataset. Is this dataset tidy? If not, why not?

This data frame is a subset (100 lines) of transcript-level gene
expression data where transcript abundance was measured at two different
time points of a certain treatment conducted in triplicates. The column
names have the format of `molecule_time_replicate`

First, explore the structure of the dataset using some of the functions
we learned in class.

```{r, q2}
```

Then, comment on whether this dataset is tidy, and if not, list the
reasons why. Hint: In a tidy dataframe, every column represents a single
variable and every row represents a single observation

```         
<your answer here>
```

### Question 3 - How will you reshape the data frame so that each row has only one experimental observation?

Hint: Use `pivot_longer()`

```{r, q3}
```

### Question 4 - How will you modify the dataframe so that multiple variables are not present in a single column?

Hint: Use `separate()`

```{r, q4}
```

### Question 5 - How will you save your output?

Hint: Use the *readr* cheatsheet to figure this out.
https://rstudio.cloud/learn/cheat-sheets

```{r, q5}
```
