---
title: "bootcamp-r-problem-set-2"
author: "Insert your name here"
date: "8/26/2020"
output: html_document
---


```{r, setup}
library(tidyverse)
library(knitr)
```

### Problem Set

Each problem below is worth **5 points**.

Use the data files in the `data/` directory to answer the questions.

For this problem set, you are allowed to help each other, but you are not allowed to post correct answers in slack.

**The problem set is due 12pm on Aug 27.**

### Grading rubric

- Everything is good: 5 points
- Partially correct answers: 3-4 points
- Reasonable attempt: 2 points

#### Question 1 - Show how you will use `arrange`, `filter`, and `select` on a dataframe

Step 1: Import the dataset `data_transcript_exp_tidy.csv` within the `data` folder using the _readr_ package
Step 2: Sort the dataframe by expression data from highest to lowest (Column: `count`) 
Step 3: Filter the dataframe by `count` > 100
Step 4: Select all columns _except_ for `type`

```{r, q1}

```

<br/>

#### Question 2 - How do you map the transcript ids to gene ids using a mapping file? 

Step 1: Import the dataset `id-mapping.csv` within the `data` folder using the _readr_ package
Step 2: Map the transcript ids in `data` to gene ids in `id-mapping.csv` 

Hint: Think `join` functions and check for extra extensions to ids that you may have to `separate` first. 
```{r, q2}

```

<br/>

#### Question 3 - How will you A) create a new column `log10count` that contains log10 transformed `count` values and B) rearrange the columns in the following order: ensembl_transcript_id, ensembl_gene_id, hgnc_symbol, type, time, replicate, count, log10count. (Note that we have dropped `extra`)

Hint: Use `mutate` and `select`
```{r, q3}

```

<br/>

#### Question 4 - How will go from transcript-level counts to gene-level counts, while keeping the `time`, `replicate`, and `hgnc_symbol` information?

Hint: Use `group_by` with multiple variables, and `summarise` the "count" values using `sum()`
```{r, q4}

```

<br/>
