---
title: "Exercises-05"
author: "Sujatha Jagannathan"
date: "8/28/2020"
---

```{r include=FALSE}
library(tidyverse)
library(knitr)
library(ggthemes) # pretty plotting themes
library(cowplot) # to make panels of plots
library(viridis) # nice colors!
library(ggridges) # for ridge plot
library(here)
```

### R bootcamp review

```{r}
# import data
data_transcript_exp_tidy <- read_csv(here("data/data_transcript_exp_tidy.csv"))
# save to have smaller name
data <- data_transcript_exp_tidy
# examples of grouping by differently - type these into the console to see the difference
group_by(data, type)
group_by(data, replicate)
group_by(data, time)
group_by(data, ensembl_transcript_id)

# example of how group_by affects the way summarise works
data |> summarise(count = mean(count, na.rm = TRUE))
group_by(data, time) |> summarise(count = mean(count, na.rm = TRUE))
group_by(data, ensembl_transcript_id) |> summarise(count = mean(count, na.rm = TRUE))
group_by(data, ensembl_transcript_id, time) |> summarise(count = mean(count, na.rm = TRUE))
```
  - functions to use within mutate - google is your friend
  - manipulation of cols & rows - use cheatsheets
  
* ggplot
  
  - aesthetic mapping: [Ref](https://ggplot2.tidyverse.org/reference/aes.html)
  
```{r, aesthetic_mapping}
# specifying colors of plots
diamonds_subset <- diamonds |> sample_n(size = 1000)

# scatter plot with color by cut - aesthetics specified in the main "mapping"
ggplot(
  data = diamonds_subset,
  mapping = aes(x = carat, y = price, color = cut)
) +
  geom_point()

# scatter plot with color by cut - aesthetics specified in the geom "mapping"
ggplot(
  data = diamonds_subset,
  mapping = aes(x = carat, y = price)
) +
  geom_point(aes(color = cut))

# color specified in the main mapping will apply universally to all geoms
ggplot(
  data = diamonds_subset,
  mapping = aes(x = carat, y = price, color = cut)
) +
  geom_point() +
  geom_smooth()

# color specified in the geom mapping only applies to that layer
ggplot(diamonds_subset, mapping = aes(x = carat, y = price)) +
  geom_point(aes(color = cut)) +
  geom_smooth()
```

  - piping data into ggplot - example below

```{r, piping_data}
diamonds_subset |> ggplot(mapping = aes(x = carat, y = price)) +
  geom_point(aes(color = cut)) +
  geom_smooth()
```

  - more applied examples - plenty to come in the rest of the course

  - specifying colors of plots: [tutorial](http://www.sthda.com/english/wiki/ggplot2-colors-how-to-change-colors-automatically-and-manually)

```{r single-color}
# coloring by a single color - more information in the tutorial above
ggplot(
  data = diamonds_subset,
  mapping = aes(x = carat, y = price)
) +
  geom_point(color = "red") +
  geom_smooth()
```
  
* other
  - log10 - `?log10`
  - How to use Help pages: help(<function>)
  - required vs. optional arguments: this is possible to distinguish if you have a well-documented function. If not, trial-an-error is how it goes. 
  - Use of commas with multiple vars: Depends on the function. Look at the exact syntax required for specific functions on the cheatsheet to know what to use: col1, col2, col3, vs col1:col3, etc
  - Metacharacters "." etc.: Refer to cheatsheet on Regex + stringr

