---
title: "R Bootcamp - Day 1"
subtitle: "R & RStudio overview"
author: "{{< var instructor.block.bootcamp-r >}}"
date: last-modified
---

```{r include=FALSE}
library(tidyverse)
library(here)
```

# Course overview 

## People

* Instructors (me, Srinivas Ramachandran, Jay Hesselberth)

* TAs (Christina Akirtava and Charlie Moffatt)

### Syllabus 

- Read the syllabus.

- Your grades are based on attendance / participation, problem sets, and a final project. Your lowest problem set grade will be dropped.

- If you are sick, let me and Srinivas know, and stay home. We will record all classes and make them available on Panopto.

## Getting help

* All course details are on the [website](https://rnabioco.github.io/molb-7950/).
* We use Canvas for problem set submission & grading.

* If you get stuck during class: use the `#class` channel in slack. TAs will come over.
* If you need help outside of class (in order):

  1. Ask a question on slack.
  2. Use calendly to schedule time with the TAs.
  3. Use calendly to schedule time with the instructors.
  4. E-mail {{< var instructor.lead >}}.

## Structure of a class

- Prior to each block (and sometimes prior to a class), check and complete material in the "Prepare" column on the class schedule.

- On the day of class and before class starts, start the day's "assignment" in Posit Cloud. This will contain blank exercises that you'll fill in during class.

- You'll also have access to the slides, but it's probably better for the first few classes to just have the exercises open.
    
## Problem sets

- You'll have a problem set assigned at the end of each class. Our expectation
is that you spend a 30-90 minutes on each problem set.

- Problem sets will get progressively more difficult.

- You can work in groups for problem sets (see the Syllabus), but during the Bootcamp
  you should avoid it.

- If you feel like you're stuck on something silly, reach out through slack or office hours.

- We'll talk about the problem set at the end of each class. You are welcome
  to use the remaining class time to start and possibly finish the problem set.

## Learning Objectives for the R Bootcamp

- Learn the fundamentals of R programming (*class 1*)

- Become familiar with "tidyverse" suite of packages

  - tidyr: "Tidy" a messy dataset (*class 2*)
  - dplyr: Transform data to derive new information (*classes 3 and 6*)
  - ggplot2: Visualize and communicate results (*classes 4 and 5*)
  - Putting all of these to use with real data sets (*classes 7 and 8*)
  
- Practice reproducible analysis using Quarto/Rmarkdown (Rigor & Reproducibility)

## Today's class outline

* Review R basics

  - R vs Rstudio (Exercises #1-2)
  - Functions & Arguments (Exercises #3-4)
  - Data types (Exercise #5)
  - Data structures (Exercises #6-7)
  - R Packages (Exercise #8)
  
* Review Quarto/Rmarkdown (Exercise #9)

## RStudio - Exercise 1

- We are using RStudio through Posit Cloud for the class.
- Look at RStudio panels one at a time.
- Environment, History, Console, Files, Plots, Packages, Help, etc.

. . .

See menu:

**Help > Cheat Sheets > RStudio IDE Cheat Sheet**

## R as a calculator - Exercise 2

- R can function like an advanced calculator

---

Try simple math.

```{r}
#| echo: true
#| output-location: fragment
# This is a comment line
# Note the order of operations (PEMDAS).
2 + 3 * 5
```

. . .

```{r}
#| echo: true
#| output-location: fragment
# value of 3-7
3 - 7
```

. . .

```{r}
#| echo: true
#| output-location: fragment
# division
3 / 2
```

. . .

```{r}
#| echo: true
#| output-location: fragment
# 5 raised to the power of 2
5^2
```

---

Assign a numeric value to an object.

- `<-` and `=` are *assignment* operators.
- By convention, R programmers use `<-`.
- `x <- 1` reads "set the value of `x` to 1".

. . .

```{r}
#| echo: true
#| output-location: fragment
# create `num` object
num <- 5^2
num
```

---

`=` and `==` are two different operators.

- a `=` is used for assignment (e.g., `x = 1`)

- a `==` tests for equivalence (e.g. `x == 1` says "does `x` equal 1?")

. . .

```{r}
#| echo: true
#| output-location: fragment
x <- 1
x == 1
```

. . .

```{r}
#| echo: true
#| output-location: fragment
x == 10
```

. . .

```{r}
#| echo: true
#| output-location: fragment
# `x` NOT equals 5?
x != 5
```

## Functions and arguments - Exercise 3

* Functions are fundamental building blocks of R
* Most functions take one or more arguments and transform an input object in a specific way. 
* Use tab-completion to find functions!

. . .

```{r}
#| echo: true
#| output-location: fragment
?log
log(4)
```

. . .

```{r}
#| echo: true
#| output-location: fragment
log(4, base = 2)
```

## Writing a simple function - Exercise 4

```{r}
#| echo: true
#| output-location: column-fragment
addtwo <- function(x) {
  num <- x + 2
  return(num)
}

addtwo(4)
```

```{r}
#| echo: true
#| output-location: column-fragment
f <- function(x, y) {
  z <- 3 * x + 4 * y
  return(z)
}

f(2, 3)
```

## Data types - Exercise 5

* There are many data types in R. 
* We'll mainly use **numeric**, **character**, and **logical**.

. . .

```{r}
#| echo: true
#| output-location: column-fragment
class(4)
```

<br>

```{r}
#| echo: true
#| output-location: column-fragment
class("jay")
```

<br>

```{r}
#| echo: true
#| output-location: column-fragment
class(TRUE)
```

<br>

```{r}
#| echo: true
#| output-location: column-fragment
# coerce one type to another
class(as.character(TRUE))
```

## Vectors - Exercise 6

Vectors are a core R data structure. 

- A vector is an ordered collection of elements of the same type (e.g. numeric, character, or logical). 

- Later you will see that every column of a data.table / tibble is a vector. 

- Operations on vectors propagate to all the elements of the vectors.

---

Let's create some vectors.

- The `c` function combines values together (e.g., `c(1,2,3)`)

. . .

```{r}
#| echo: true
#| output-location: column-fragment
x <- c(1, 3, 2, 10, 5)
x
```

. . .

```{r}
#| echo: true
#| output-location: column-fragment
class(x)
```

---

```{r}
#| echo: true
#| output-location: column-fragment
y <- 1:5

y + 2
```

<br>

```{r}
#| echo: true
#| output-location: column-fragment
2 * y
```

<br>

```{r}
#| echo: true
#| output-location: column-fragment
y^2
```

---

```{r}
#| echo: true
#| output-location: column-fragment
# `y` has not changed!
y
```

<br>

```{r}
#| echo: true
#| output-location: column-fragment
# this will update the value of `y`
y <- y * 2
y
```

## Data frames

- A data.frame is a rectangle, where each column is a vector, and each row is a slice across vectors.

- data.frame columns are vectors, and can have different types (numeric, character, factor, etc.). 

- A data.frame is constructed with `data.frame()`.

---

```{r}
#| echo: true
#| output-location: fragment
class(iris)
```

<br>

```{r}
#| echo: true
#| eval: false
iris
```

<br>

```{r}
#| echo: true
#| output-location: fragment
data.frame(x = c(1, 2, 3), y = c(2, 4, 6))
```

## Tibbles

* A *tibble* is a modern reimagining of the data.frame, keeping what time has proven to be effective, and throwing out what is not. 

* Tibbles are data.frames that are lazy and surly: they do less (i.e. they don't change variable names or types, and don't do partial matching) and complain more (e.g. when a variable does not exist). 

* This forces you to confront problems earlier, typically leading to cleaner, more expressive code. Tibbles also have an enhanced `print()` method which makes them easier to use with large datasets containing complex objects.

---

* `tibble()` does much less than `data.frame()`: 

  - it never changes the type of the inputs 
  - it never changes the names of variables
  - it never creates `row.names()`

## Data frames & tibbles - Exercise 7

Create a data.frame and tibble.

```{r}
#| echo: true
chrom <- c("chr1", "chr1", "chr2")
start <- c(200, 4000, 100)
end <- c(250, 410, 200)
strand <- c("-", "-", "+")

df <- data.frame(chrom, start, end, strand)

tbl <- tibble(chrom, start, end, strand)
```

. . .

Now echo the contents of `df` and `tbl` to the console and inspect

## R packages - Exercise 8

* An R package is a collection of code, data, documentation, and tests that is easily shareable.

* A package often has a collection of custom functions that enable you to carry out a workflow. eg. DESeq for RNA-seq analysis.

* The most popular places to get R packages from are CRAN, Bioconductor, and Github. 

* Once a package is installed, one still has to "load" them into the environment using a `library(<package>)` call. 

---

Let's do the following to explore R packages:

* Look at the "Environment" panel in Rstudio
* Explore Global Environment
* Explore the contents of a package

## Quarto Exercise - Exercise 9

* [Quarto](https://quarto.org) is a fully reproducible authoring framework to create, collaborate, and communicate your work. 

* Quarto lets you render Rmarkdown documents (in addition to Jupyter notebooks, etc.)

* Quarto supports a number of output formats including pdfs, word documents, slide shows, html, etc. 

---

* A Quarto document is a plain text file with the extension `.qmd` and contains the following basic components: 

  - A YAML header surrounded by `---`.
  - Chunks of R code surrounded by ```.
  - Plain text structured with [markdown](https://www.markdownguide.org/basic-syntax) formatting like `# heading` and `*italics*`.

---

Let's do the following to explore Quarto documents:

* Create a new Quarto document
* Render the document to see the output

## Problem sets and submission

Your first problem set is in `problem-sets/class-01.qmd`

