# Block 4: Protein {#block-protein}


This 3 week section covers specific experiments performed on proteins and the types of molecular insights they can provide. 

## Week 7: Mass Spectrometry

Mass spectrometry (MS) is a powerful technique used to identify and quantify proteins by analyzing peptides generated from samples. We will cover the principles, variety of methods, and important data analysis considerations of MS experiments (figure from Wikipedia).  


```{r fig.width=6, fig.height=4, fig.align='center', echo=FALSE}
library(png)
library(grid)
img <- readPNG("classes/images/Mass_spectrometry.png")
grid.raster(img)
```

## Week 8: Densitometry

A recurrent principle in molecular biology is to seprate molecules by molecular mass using gel electrophoresis, more specifically Western bloting for protein analsis. This is very common and we will cover methods to perform analysis and quantification of these experiments using densitometry methods (figure from Wikipedia).

```{r fig.width=6, fig.height=4, fig.align='center', echo=FALSE}
library(png)
library(grid)
img <- readPNG("classes/images/western.png")
grid.raster(img)
```


## Week 9: Immunofluoresence

Determining the precise location of a specific protein either within a cell or organ provides critical insights. We will cover important considerations and common pitfalls for the analysis of protein localization by immunofluoresence experiments (figure from Wikipedia). 


```{r fig.width=6, fig.height=4, fig.align='center', echo=FALSE}
library(png)
library(grid)
img <- readPNG("classes/images/IF.png")
grid.raster(img)
```
