---
title: "R Bootcamp - Day 5"
subtitle: "more ggplot2"
author: "Jay Hesselberth"
---

```{r}
#| include: false
library(tidyverse)
library(knitr)
library(here)

library(cowplot) # to make panels of plots
library(ggridges) # ridge plots
```

## Class 4 & 5 outline {.smaller}

:::: {.columns}

::: {.column width="50%"}
- Introduce ggplot2 & today's data sets (Exercise 1)
- Understand the basics of ggplot2 (Exercise 2, 3)
- Geom functions (Exercise 4-8)
- Geom_point properties (Exercise 9)
- Position adjustments (Exercise 10)
- Coordinate and Scale Functions (Exercise 11)
- Zooming into a plot (Exercise 12)
:::

::: {.column width="50%"}
- Faceting (Exercise 13)
- Themes (Exercise 14)
- Labels & Legends (Exercise 15)
- Adding lines to plots (Exercise 16)
- Making multi-panel figures (Exercise 17)
- Saving a plot (Exercise 18)
:::

::::

## shape, size, fill, color, and transparency - Exercise 9

Get a diamonds subset.

```{r}
```

Note that aesthetics can also be defined *within* a geom.

This is useful if you use two different geoms that share an aesthetic.

```{r}
```

## Position adjustments - Exercise 10

A stacked bar chart.

```{r}
```

Dodged bars are easier to read (proportions are clearer)

```{r}
```

## Coordinate and Scale Functions - Exercise 11

Logarithmic axes - 1

Note the difference between axis labels in these two examples.

```{r}
```

---

Logarithmic axes - 2

```{r}
```

---

Flipping coordinate system (swapping `x` and `y`)

```{r}
```

---

Now flip the axis.

```{r}
```

Brief aside: ggplot can handle on-the-fly data transformations.

Here we log-transform carat and convert USD to CAD.

```{r}
```

## Zooming into a plot - Exercise 12

We might want to change the limits of x or y axes to zoom in.

```{r}
```

## Faceting to plot subsets of data into separate panels - Exercise 13

A density plot we've seen before.

**Which variables can we use to subdivide the data?**

```{r}
```

---

Faceted by `cut`

```{r}
```

Let's also use `facet_grid()` to facet by two variables.

Faceted by `clarity` and `cut`.

```{r}
```

---

Scatter plot with facets.

```{r}
```

## Themes - Exercise 14

Scatter plot with default theme.

```{r}
```

Change the theme with `theme_bw()`.

```{r}
```

**My go-to is `cowplot::theme_cowplot()`.**

It implements much of the advice in the "Dataviz" book, i.e.. YOUR LABELS ARE TOO
SMALL.

```{r}
```

We're not going to cover it, but you can also customize pre-existing themes.

## Labels & Legends - Exercise 15

Use `labs()` to add / change plot labels.

```{r}
```

# *Additional points*

## How to add a line to a plot? (Exercise 16)

```{r}
```

---

Also try:

```{r}
```

## How to combine multiple plots into a figure? (Exercise 17)

```{r}
```

We have 4 legends - can they be condensed?

Yes, but it is not exactly straightforward.

**need to scroll below**

```{r}
```

## Saving plots (Exercise 18)

Saves last plot as 5' x 5' file named "plot_final.png" in working directory.

Matches file type to file extension.

```{r}
```
