---
title: "R Bootcamp - Day 4"
subtitle: "ggplot2"
author: "Jay Hesselberth"
---

```{r}
#| include: false
library(tidyverse)
library(knitr)
library(here)

library(cowplot) # to make panels of plots
library(ggridges) # ridge plots
```

## *ggplot2*

## Today's datasets

In this class, we will use a data set from ggplot2: `diamonds` contains thousands of gem prices and qualities.

There are many interesting data sets you can install as R packages for learning to manipulate and plot data:

- [babynames](https://hadley.github.io/babynames/)
- [gapminder](https://jennybc.github.io/gapminder/)
- [palmerpenguins](https://allisonhorst.github.io/palmerpenguins/)

## Getting familiar with the data - Exercise 1 {.smaller}

```{r}
```

## The syntax of `ggplot()`  

## Making a plot step-by-step (Exercise 2)

---

Initialize a plot with data.

```{r}
```

Next, specify the coordinate system.

```{r}
```

Add a geom (`geom_point`).

```{r}
```

Map aesthetics to other variables.

```{r}
```

---

Reduce overplotting by adjusting the transparency of points.

```{r}
```

## Looking under the hood of ggplot (Exercise 3) {.smaller}

```{r}
```

## ggplot is powerfully simple for making complex plots

```{r}
```

Why can't I just do this?

```{r}
```

## Geom functions

* Use a geom function to represent data points, use the geom aesthetic properties to represent variables.
* Each function returns a plot layer.
* There are many geoms in ggplot that are specific to plots with 1, 2, or 3 variables 

Make a bar plot.

```{r}
```

---

Update the bar plot aesthetics.

```{r}
```

---

Change to a density plot.

```{r}
```

---

Color the density plot.

```{r}
```

---

Plot subsets by mapping `fill` to `cut`

```{r}
```

---

Use `ggridges` to plot staggered subsets.

https://wilkelab.org/ggridges/

```{r}
```

## Geom functions for two variables

Make a column plot.

```{r}
```

Same data with a box plot.

```{r}
```

---

Box plot, with fill color by cut.

```{r}
```

Violin plot with fill color by cut.

```{r}
```

## continuous x, continuous y - Exercise 6

Subset diamonds to see points more clearly.

```{r}
```

Make a scatter plot.

```{r}
```

Now add a smoothing line.

```{r}
```

Here we can combine geoms to see points & the fit

```{r}
```
