---
title: "R Bootcamp - Day 1 - Exercises"
author: "Jay Hesselberth"
date: last-modified
---

## Exercise overview 

* R, Rstudio (Exercise #1)
* R as a calculator (Exercise #2)
* Functions and arguments (Exercises #3-4)
* Data types: numeric, character, logical (& more) (Exercise #5)
* Data structures: vector, list, matrix, data frame, tibbles (Exercises #6-7)
* Package system, Rstudio, and Rmarkdown (Exercises #8-9)

### R vs Rstudio - Exercise 1

* Look at Rstudio panels one at a time.
* Environment, History, Console, Files, Plots, Packages, Help, etc. 

### R as a calculator - Exercise 2

Try simple math.

```{r}

```

Assign a numeric value to a variable.

```{r}

```

Print the object to check.

```{r}

```

Compute with the variable.

```{r}

```

### Functions and arguments - Exercise 3

```{r}

```

### Writing a simple function - Exercise 4

```{r}

```

### Data types - Exercise 5

```{r}
```

### Vectors - Exercise 6

Create the following vectors:

```{r}

```

### Data frames - Exercise 7

```{r}

```

### R packages - Exercise 8

* Look at the "Environment" panel in Rstudio
* Explore Global Environment
* Explore the contents of a package

### Rmarkdown Exercise - Exercise 9

Let's do the following to explore Rmarkdown documents:

* Create a new `.Rmd` document (File > New File > Rmarkdown)
* `knit` the document to see the output

### Homework instructions

* Today's homework: 

1. Go over everything we covered today and make sure you understand it. (Use office hours if you have questions) - Expected time spent: 30 min - 1 hour

2. Go over Rstudio and Rmarkdown cheatsheets (Finding cheatsheets: Exercise 10) - - Expected time spent: 30 min on each cheatsheet
 