---
title: "R Bootcamp - Day 2 - Exercises"
author: "Jay Hesselberth"
date: last-modified
---

```{r include=FALSE}
library(tidyverse)
library(knitr)
library(here)
```

# Tidying exercises

## Data sets for tidying - Exercise 2

* Explore `table1`, `table2`, `table3`, `table4a`, `table4b`, and `table5`,
which all display the number of TB cases documented by the World Health
Organization in Afghanistan, Brazil, and China between 1999 and 2000. 

```{r}
```

## Getting familiar with the data - Exercise 3

R provides many functions to examine features of a data object

  - `View()` - To open the table up in an excel-like interface - not recommended for large tables
  - `class()` - what kind of object is it (high-level)?
  - `typeof()` - what is the object’s data type (low-level)?
  - `is_tibble()` - use is.? to confirm data type
  - `str()` - what is the structure of the object?
  - `attributes()` - does it have any metadata?

* Let's explore `table1`

```{r}
```

## Getting familiar with the data - Exercise 4

Some of the useful functions for data.frames / tibbles are as follows:  

- `head()` - shows first 6 rows  
- `tail()` - shows last 6 rows  
- `dim()` - returns the dimensions of data frame (i.e. number of rows and number of columns)  
- `nrow()` - number of rows  
- `ncol()` - number of columns  
- `names()` or `colnames()` - both show the names attribute for a data frame  
- `glimpse()`

```{r}
```

# Exercise 5 -Getting familiar with data - _summary_, _hist_, & _table_

_summary_: A generic function used to produce result summaries of the results of various model fitting functions.

```{r}
```

_hist_: Takes in a vector of values and plots a histogram.

```{r fig.height= 4, fig.width= 4}
```

_table_: Uses the cross-classifying factors to build a contingency table of the counts at each combination of factor levels.

```{r}
```

## pivot_wider - Exercise 6

What will the output look like?

```{r}
```

If you want to save the output, assign it to a new variable. This new variable
will appear in your Environment tab.

```{r}
```

## pivot_longer - Exercise 7

What will the output look like?

```{r}
```

## separate - Exercise 8

What will the output look like?

```{r}
```

### separate_rows - Exercise 9

```{r}
```

## unite - Exercise 10

```{r}
```

## Missing values

```{r}
```
