---
title: "R Bootcamp - Day 1"
subtitle: "R & RStudio overview"
author: "Jay Hesselberth"
date: last-modified
---

{{< include ./_extensions/r-wasm/live/_knitr.qmd >}}

```{r}
#| include: false
library(tidyverse)
library(knitr)
library(here)
```

## RStudio - Exercise 1

- We are using RStudio through Posit Cloud for the class.
- Look at RStudio panels one at a time.
- Environment, History, Console, Files, Plots, Packages, Help, etc.

See menu:

**Help > Cheat Sheets > RStudio IDE Cheat Sheet**

## R as a calculator - Exercise 2

- R can function like an advanced calculator

---

Try simple math.

```{webr}
2 + 1
```


Assign a numeric value to an object.

- `<-` and `=` are *assignment* operators.
- By convention, R programmers use `<-`.
- `x <- 1` reads "set the value of `x` to 1".

. . .

```{r}
```


`=` and `==` are two different operators.

- a `=` is used for assignment (e.g., `x = 1`)

- a `==` tests for equivalence (e.g. `x == 1` says "does `x` equal 1?")


```{r}
```


```{r}
```


```{r}
```

## Functions and arguments - Exercise 3

```{r}
```

```{r}
```

## Writing a simple function - Exercise 4

```{r}
```

```{r}
```

## Data types - Exercise 5

* There are many data types in R.
* We'll mainly use **numeric**, **character**, and **logical**.

```{r}
```


```{r}
```


```{r}
```

```{r}
```

## Vectors - Exercise 6

Let's create some vectors.

- The `c` function combines values together (e.g., `c(1,2,3)`)


```{r}
```

. . .

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```

## Data frames

- A data.frame is a rectangle, where each column is a vector, and each row is a slice across vectors.

- data.frame columns are vectors, and can have different types (numeric, character, factor, etc.).

- A data.frame is constructed with `data.frame()`.

```{r}
```

```{r}
```

```{r}
```

## Data frames & tibbles - Exercise 7

Create a data.frame and tibble.

```{r}
```

Now echo the contents of `df` and `tbl` to the console and inspect

## R packages - Exercise 8

Let's do the following to explore R packages:

* Look at the "Environment" panel in Rstudio
* Explore Global Environment
* Explore the contents of a package

## Quarto Exercise - Exercise 9

Let's do the following to explore Quarto documents:

* Create a new Quarto document
* Render the document to see the output

## Problem sets and submission

Your first problem set is in `problem-sets/ps-01.qmd`
