---
title: "Stats Bootcamp - class 10"
subtitle: "Stats intro and history"
author: "Neelanjan Mukherjee"
editor: visual
---

```{r}
#| echo: false
#| include: false
library(tidyverse)
library(cowplot)
```

## Let's flip a fair coin {.smaller}

Flip a coin 5 times with equal prob of H or T

```{r}
#| echo: true
#| output-location: column-fragment



```

Again

```{r}
#| echo: true
#| output-location: column-fragment



```

. . .

Set a seed

```{r}
#| echo: true
#| output-location: column-fragment

# 


```

```{r}
#| echo: true
#| output-location: column-fragment

# 


```

## Let's flip an unfair coin {.smaller}

Flip a coin 5 times with equal prob of H or T

```{r}
#| echo: true
#| output-location: column-fragment


```

Again

```{r}
#| echo: true
#| output-location: column-fragment


```

## Let's summarize the flipping results {.smaller}

Flip a fair coin 10

```{r}
#| echo: true
#| output-location: column-fragment


```

Flip a fair coin 10 and calculate mean

```{r}
#| echo: true
#| output-location: column-fragment



```

Again

```{r}
#| echo: true
#| output-location: column-fragment



```

. . .

Unfair coin

```{r}
#| echo: true
#| output-location: column-fragment



```

Unfair coin, again

```{r}
#| echo: true
#| output-location: column-fragment



```

## Let's go wild flipping {.smaller}

Flip a fair coin 10 times and calculate mean. Then do 5 rounds of that experiment.

```{r}
#| echo: true
#| output-location: column-fragment


```

. . .

Same thing for an unfair coin.

```{r}
#| echo: true
#| output-location: column-fragment


```

## Tidy and visualize flips {.smaller}

make a dataframe with means and accompanying info

```{r}
#| echo: true
#| output-location: column-fragment


```

. . .

plot it

```{r}
#| echo: true
#| output-location: column-fragment


```

## Play around some more {.smaller}

```{r}
#| echo: true
#| output-location: column-fragment


```

