---
title: "R Bootcamp Problem Set 3"
author: "Insert your name here"
---

## Setup

Start by loading libraries you need analysis in the code chunk below.

When in doubt, start by loading the tidyverse package.

```{r, setup}
library(tidyverse)
library(here)
```

## Problem Set

Each problem below is worth **5 points**.

Use the data files in the `data/` directory to answer the questions.

For this problem set, you are allowed to help each other, but you are not allowed to post correct answers in slack.

**The problem set is due 12pm on Aug 31.**

### Grading rubric

- Everything is good: 5 points
- Partially correct answers: 3-4 points
- Reasonable attempt: 2 points

## Question 1 

Use `arrange`, `filter`, and `select` on a tibble. Do the following, in order:

1. Import the data set `data/data_transcript_exp_tidy.csv.gz` (use the `here` package). 
2. Sort the tibble by expression data (`count`) from highest to lowest level.
3. Filter the tibble by `count` > 100
4. Select all columns *except* for `type`
5. Assign the output to a new object `exp_tbl`.

```{r, q1}
raw_tbl <- read_csv(
  here("data/data_transcript_exp_tidy.csv.gz")
)

exp_tbl <- raw_tbl |> 
  arrange(desc(count)) |>
  filter(count > 100) |>
  select(-type)

exp_tbl
```


## Question 2

Next, do some manipulation on `exp_tbl`:

1. Create a new variable `count_log` that contains `log10`-transformed `count` values

```{r, q3}
exp_tbl |>
  mutate(count_log = log10(count))
```

## Question 3 

Nowtranscript-level counts to gene-level counts, while keeping the `time`, `replicate`, and `hgnc_symbol` information?**

Hint: Use `group_by` with multiple variables, and `summarise` the "count" values using `sum()`

```{r, q4}

```
