---
title: "Problem Set Stats Bootcamp - class 13"
subtitle: "Hypothesis Testing"
author: "Neelanjan Mukherjee"
editor: visual
---

```{r, echo=FALSE}
library(tidyverse)
library(rstatix)
library(ggpubr)
library(gt)
library(janitor)
library(here)
library(broom)
```

```{r}
biochem <- read_tsv("http://mtweb.cs.ucl.ac.uk/HSMICE/PHENOTYPES/Biochemistry.txt", show_col_types = FALSE) |>
  janitor::clean_names()

# simplify names a bit more
colnames(biochem) <- gsub(pattern = "biochem_", replacement = "", colnames(biochem))

# we are going to simplify this a bit and only keep some columns
keep <- colnames(biochem)[c(1, 6, 9, 14, 15, 24:28)]
biochem <- biochem[, keep]

# get weights for each individual mouse
# careful: did not come with column names
weight <- read_tsv("http://mtweb.cs.ucl.ac.uk/HSMICE/PHENOTYPES/weight", col_names = F, show_col_types = FALSE)

# add column names
colnames(weight) <- c("subject_name", "weight")

# add weight to biochem table and get rid of NAs
# rename gender to sex
b <- inner_join(biochem, weight, by = "subject_name") |>
  na.omit() |>
  rename(sex = gender)
```

## Problem \# 1

Is there an association between mouse calcium and sodium levels?

### 1. Make a scatterplot to inspect variable --- (2 pts)

```{r}
ggscatter(
  data = b,
  y = "calcium",
  x = "sodium"
)
```

### 2. Are they normal (enough)? --- (1 pts)

```{r}
ggqqplot(data = b, x = "calcium")

b |> shapiro_test(calcium)

ggqqplot(data = b, x = "sodium")

b |> shapiro_test(sodium)
```

Which test will you use and why?

> I will use the Pearson since the qqplots look ok and there are \~1700 observations. Spearman is fine too - especially since sodium looks a little like an integer and not continuous.

### 3. Declare null hypothesis $\mathcal{H}_0$ --- (1 pts)

$\mathcal{H}_0$ is that there is no dependency/association between $calcium$ and $sodium$

### 4. Calculate and plot the correlation on a scatterplot --- (2 pts)

```{r}
ggscatter(
  data = b,
  y = "calcium",
  x = "sodium"
) +
  stat_cor(
    method = "pearson",
    label.x = 110,
    label.y = 2.4
  )
```

## Problem \# 2

## Do mouse calcium levels explain mouse sodium levels? If so, to what extent?

Use a linear model to do the following:

### 1. Specify the Response and Explanatory variables --- (2 pts)

> The response variable y is sodium The explantory variable x is calcium

### 2. Declare the null hypothesis --- (1 pts)

> The null hypothesis is calcium levels do not explain/predict sodium levels.

### 3. Use the `lm` function to create a fit (linear model) --- (1 pts)

also save the slope and intercept for later

```{r lm fit}
fit <- lm(formula = sodium ~ 1 + calcium, data = b)


int <- tidy(fit)[1, 2]
slope <- tidy(fit)[2, 2]
```

### 4. Add residuals to the data and create a plot visualizing the residuals --- (1 pts)

```{r augment and viz}
b_fit <- augment(fit, data = b)

avg_sod <- mean(b$sodium)

ggplot(
  data = b_fit,
  aes(x = calcium, y = sodium)
) +
  geom_point(size = 1, aes(color = .resid)) +
  geom_abline(
    intercept = pull(int),
    slope = pull(slope),
    col = "red"
  ) +
  scale_color_gradient2(
    low = "blue",
    mid = "black",
    high = "yellow"
  ) +
  geom_segment(
    aes(
      xend = calcium,
      yend = .fitted
    ),
    alpha = .1
  ) + # plot line representing residuals
  theme_linedraw()
```

### 5. Calculate the $R^2$ and compare to $R^2$ from fit --- (2 pts)

$R^2 = 1 - \displaystyle \frac {SS_{fit}}{SS_{null}}$

$SS_{fit} = \sum_{i=1}^{n} (data - line)^2 = \sum_{i=1}^{n} (y_{i} - (\beta_0 \cdot 1+ \beta_1 \cdot x)^2$

$SS_{null}$ --- sum of squared errors around the mean of $y$

$SS_{null} = \sum_{i=1}^{n} (data - mean)^2 = \sum_{i=1}^{n} (y_{i} - \overline{y})^2$

```{r rsq}
ss_fit <- sum(b_fit$.resid^2)
ss_null <- sum(
  (b_fit$sodium - avg_sod)^2
)
rsq <- 1 - ss_fit / ss_null
glance(fit) |> select(r.squared)
```

### 6. Using $R^2$, describe the extent to which calcium explains sodium levels --- (2 pts)

> $calcium$ explains \~68% of variation in $sodium$ levels

### 7. Report (do not calculate) the $p-value$ and your decision on the null --- (1 pts)

```{r }
fit |>
  glance() |>
  select(p.value)
```

> The null hypothesis is calcium levels do not explain/predict sodium levels IS NOT SUPPORTED

Calcium levels to predict sodium levels.

## Problem \# 3

What is the association between mouse weight and age levels for different sexes?

### 1. Calculate the pearson correlation coefficient between weight and age for females and males --- (2 pts)

```{r}
b |>
  group_by(factor(sex)) |>
  cor_test(weight, age)
```

### 2. Describe your observations --- (2 pts)

> The relationship between weight and age is stronger for males than it is for females.
