---
title: "Problem Set Stats Bootcamp - class 14"
subtitle: "Hypothesis Testing"
author: "Neelanjan Mukherjee"
editor: visual
---

```{r}
#| echo: false
library(tidyverse)
library(rstatix)
library(ggpubr)
library(gt)
library(janitor)
library(here)
library(cowplot)
library(broom)
```

```{r}
biochem <- read_tsv("http://mtweb.cs.ucl.ac.uk/HSMICE/PHENOTYPES/Biochemistry.txt", show_col_types = FALSE) |>
  janitor::clean_names()

# simplify names a bit more
colnames(biochem) <- gsub(pattern = "biochem_", replacement = "", colnames(biochem))

# we are going to simplify this a bit and only keep some columns
keep <- colnames(biochem)[c(1, 6, 9, 14, 15, 24:28)]
biochem <- biochem[, keep]

# get weights for each individual mouse
# careful: did not come with column names
weight <- read_tsv("http://mtweb.cs.ucl.ac.uk/HSMICE/PHENOTYPES/weight", col_names = F, show_col_types = FALSE)

# add column names
colnames(weight) <- c("subject_name", "weight")

# add weight to biochem table and get rid of NAs
# rename gender to sex
b <- inner_join(biochem, weight, by = "subject_name") |>
  na.omit() |>
  rename(sex = gender)
```

## Problem \# 1

Can mouse sex explain mouse cholesterol? {.smaller}

## STEP 1: Null hypothesis and variable specification --- (2 pts)

$\mathcal{H}_0:$ mouse $sex$ does NOT explain $cholesterol$

> $cholesterol$ is the response variable

> $sex$ is the explanatory variable

## STEP 2: Fit linear model and examine results --- (3 pts) {.smaller}

```{r}
#| echo: true

fit_cs <- lm(formula = tot_cholesterol ~ 1 + sex, data = b)
```

Fit summary:

```{r}
#| echo: false
#| echo: true
#| output-location: column-fragment
glance(fit_cs) |>
  gt() |>
  fmt_number(columns = r.squared:statistic, decimals = 3)
```

Coefficient summary:

```{r}
#| echo: false
#| echo: true
#| output-location: column-fragment
tidy(fit_cs) |>
  gt() |>
  fmt_number(columns = estimate:statistic, decimals = 3)
```

## Collecting residuals and other information --- (2 pts) {.smaller}

add residuals and other information

```{r}
#| echo: true
#| output-location: column-fragment


# augment
b_cs <- augment(fit_cs, data = b)


avg_c <- mean(b_cs$tot_cholesterol)

c <- b |>
  group_by(sex) |>
  get_summary_stats(tot_cholesterol, type = "mean")


# mean chol female
avg_cf <- pull(c[1, 4])


# mean chol male
avg_cm <- pull(c[2, 4])
```

## STEP 4: Visualize the error around fit --- (2 pts) {.smaller}

```{r}
#| echo: false
#| echo: true
#| output-location: column-fragment
#| fig.alt: "Description of the plot - PLEASE FILL IN"


# plot of data with mean and colored by residuals
p_cs <- ggplot(
  b_cs,
  aes(x = sex, y = tot_cholesterol)
) +
  geom_point(
    position = position_jitter(),
    aes(color = .resid)
  ) +
  scale_color_gradient2(
    low = "blue",
    mid = "black",
    high = "yellow"
  ) +
  geom_hline(
    yintercept = avg_c,
    color = "darkgrey"
  ) +
  geom_segment(
    aes(
      x = .5, xend = 1.5,
      y = avg_cf, yend = avg_cf
    ),
    color = "red"
  ) +
  geom_segment(
    aes(
      x = 1.5, xend = 2.5,
      y = avg_cm
    ),
    yend = avg_cm,
    color = "red"
  ) +
  theme_cowplot()

p_cs
```

## STEP 5: Visualize the error around the null (mean weight) --- (2 pts) {.smaller}

```{r}
#| echo: false
#| echo: true
#| output-location: column-fragment
#| fig.alt: "Description of the plot - PLEASE FILL IN"

p_c <- ggplot(
  b_cs,
  aes(x = sex, y = tot_cholesterol)
) +
  geom_point(
    position = position_jitter(),
    aes(color = tot_cholesterol - avg_c)
  ) +
  scale_color_gradient2(
    low = "blue",
    mid = "black",
    high = "yellow"
  ) +
  geom_hline(
    yintercept = avg_c,
    color = "darkgrey"
  ) +
  theme_cowplot()

p_c
```

## Plot the fit error and the null error as 2 panels --- (2 pts) {.smaller}

```{r}
#| echo: true
#| fig.alt: "Description of the plot - PLEASE FILL IN"

plot_grid(p_cs, p_c, ncol = 2, labels = c("cholesterol by sex", "cholesterol"))
```

## Calculate $R^2$ --- (3 pts) {.smaller}

```{r}
#| echo: true
#| output-location: column-fragment

ss_fit <- sum(b_cs$.resid^2)

ss_null <- sum(
  (b_cs$tot_cholesterol - avg_c)^2
)
```

$R^2 = 1 - \displaystyle \frac {SS_{fit}}{SS_{null}}$

```{r}
#| echo: true
#| output-location: column-fragment

rsq <- 1 - ss_fit / ss_null
rsq
```

check agains Rsq in your fit

```{r}
#| echo: true
#| output-location: column-fragment

glance(fit_cs) |> select(r.squared)
```

## Compare to traditional t-test --- (2 pts) {.smaller}

```{r}
#| echo: true
#| output-location: column-fragment

# run analagous t-test
b |>
  t_test(tot_cholesterol ~ 1 + sex) |>
  select(-c(n1, n2, df)) |>
  gt()
```

```{r}
#| echo: true
#| output-location: column-fragment

tidy(fit_cs) |>
  select(term, estimate, statistic, p.value) |>
  gt()
```

## Provide your interpretation of the result --- (2 pts)

The null model that mouse $sex$ does NOT explain $cholesterol$ is not well supported. Therefore, i believe that mouse $sex$ is able to explain \~%18 of the variation in $cholesterol$.
