---
title: "Problem Set 2 Key"
author: "JH"
---

```{r}
#| label: setup
library(tidyverse)
library(here)
```

## Problem Set

Each problem below is worth **4 points**.

Use the data files in the `data/` directory to answer the questions.

For this problem set, you are allowed to help each other, but you are
not allowed to post correct answers in slack.

**The problem set is due 12pm on Aug 30.**

### Grading rubric

- Everything is good: 5 points
- Partially correct answers: 3-4 points
- Reasonable attempt: 2 points

## Question 1 

Import the dataset `data_transcript_exp_subset` using the *readr*
package.

Hint: The file is located at the following path
`data/data_transcript_exp_subset.csv.gz`

```{r}
#| label: q1
x <- read_csv(here("data/data_transcript_exp_subset.csv.gz"))
```

## Question 2

**Explore the dataset. Is this dataset tidy? If not, why not?**

This data frame is a subset (100 lines) of transcript-level gene
expression data where transcript abundance was measured at two different
time points of a certain treatment conducted in triplicates. The column
names have the format of `molecule_time_replicate`

First, explore the structure of the dataset using some of the functions
we learned in class.

```{r}
#| label: q2
x

summary(x)

glimpse(x)
```

Comment on whether this dataset is tidy, and if not, list the
reasons why. Hint: In a tidy dataframe, every column represents a single
variable and every row represents a single observation

### Answer

It is not tidy because the time points and replicates are not in their own columns.

## Question 3 

**How will you reshape the data frame so that each row has only one experimental observation?**

Hint: Use `pivot_longer()`

```{r}
#| label: q3
x |> pivot_longer(-ensembl_transcript_id)
```

### Question 4

**How will you modify the dataframe so that multiple variables are not present in a single column?**

Hint: Use `separate()`

```{r}
#| label: q4
x_tidy <- x |>
  pivot_longer(-ensembl_transcript_id) |>
  separate(name, into = c("mol", "time", "rep"), sep = "_")

x_tidy
```

### Question 5 

**How will you save your output as a TSV file?**

Hint: Use the *readr* cheatsheet to figure this out.

https://rstudio.cloud/learn/cheat-sheets

```{r}
#| label: q5
write_csv(x_tidy, "transcripts.tidy.csv")
```
