---
title: "Problem Set 1 Key"
author: "JH"
---

## Problem Set

Each problem below is worth **10 points**.

**The problem set is due 12pm on Aug 30.**

### Grading rubric

- Everything is good: 5 points
- Partially correct answers: 3-4 points
- Reasonable attempt: 2 points

## Setup

Start by loading libraries you need analysis below. When in doubt, start by loading the tidyverse package.

```{r}

```

## Question 1

Create 3 different vectors called `x`, `y`, and `z`:

- `x` should be character vector of length 5
- `y` should be a numerica vector of length 5
- `z` should be a logical vector of length 5

Use `length()` to calculate the length of each vector.

```{r}
x <- LETTERS[1:5]
y <- 1:5
z <- c(TRUE, TRUE, FALSE, FALSE, FALSE)

x
y
z

length(x)
length(y)
length(z)
```

## Question 2

Using the vectors you created above, create a new tibble with column names
`x`, `y`, and `z`.

Use `nrow()` and `ncol()` to calculate the number of rows and columns.

What do you notice about the length of the vectors and the number of rows.

```{r}
tbl <- tibble(x = x, y = y, z = z)
nrow(tbl)
ncol(tbl)
```

### Answer

The length of the vectors and the number of rows are the same, because
tibble columns are simply the vectors we started with.

## Submit

Be sure to click the "Render" button to render the HTML output.

Then paste the URL of this Posit Cloud project into the problemt set on Canvas.
