---
title: "Problem Set Stats Bootcamp - class 12"
subtitle: "Hypothesis Testing"
author: "Neelanjan Mukherjee"
editor: visual
---

```{r, echo=FALSE}
library(tidyverse)
library(rstatix)
library(ggpubr)
library(gt)
library(janitor)
library(here)
```

## Problem \# 1

Does mouse sex explain mouse total cholesterol levels? Make sure to run chunks above.

### 1. Examine and specify the variable(s) (1 pt)

> The response variable y is $tot\_cholesterol$\
> The explantory variable x is $sex$

### Make a violin plot: (2 pt)

response variable on the y-axis

explanatory variable on the x-axis

```{r, eval = FALSE}
ggviolin(
  data = b,
  y = "tot_cholesterol",
  x = "sex",
  fill = "sex",
  add = "mean_sd"
)
```

### Get n, mean, median, sd (1 pt)

```{r, eval = FALSE}
b |>
  group_by(sex) |>
  get_summary_stats(tot_cholesterol,
    type = "common",
    show = c("mean", "median", "sd")
  )
```

### Is it normally distribute? (1 pt)

```{r, eval = FALSE}
ggqqplot(
  data = b,
  x = "tot_cholesterol",
  color = "sex"
)


b |>
  group_by(sex) |>
  shapiro_test(tot_cholesterol) |>
  gt()
```

> Yes, based on the qq-plot and the high $n$, but i do understand if you want to play it safe due to the shapiro_test p-value.

### Is it variance similar between groups? (1 pt)

```{r, eval = FALSE}
b |>
  levene_test(tot_cholesterol ~ sex) |>
  gt()
```

> Yes

### What kind of test are you picking and why? (1 pt)

> t_test since i think it is normally distribute, with equal variance based on levene test

### 2. Declare null hypothesis $\mathcal{H}_0$ (1 pt)

$\mathcal{H}_0$ is that $sex$ does not explain $tot\_cholesterol$

### 3. Calculate test-statistic, exact p-value and plot (2 pt)

```{r, eval = FALSE}
b |>
  t_test(tot_cholesterol ~ sex,
    var.equal = T,
    ref.group = "F"
  ) |>
  gt()

statres <- b |>
  t_test(tot_cholesterol ~ sex,
    var.equal = T,
    ref.group = "F"
  )


ggviolin(
  data = b,
  y = "tot_cholesterol",
  x = "sex",
  fill = "sex",
  add = "mean_sd"
) +
  stat_pvalue_manual(
    statres,
    label = "p",
    y.position = 5.8
  ) +
  ylim(1, 6)
```

> can reject null hypothesis

## Problem \# 2

Does mouse family explain mouse total cholesterol levels? Make sure to run chunk above.

### 1. Examine and specify the variable(s) (1 pt)

> The response variable y is $tot\_cholesterol$\
> The explantory variable x is $family$

### Make a plot: (2 pt)

response variable on the y-axis

explanatory variable on the x-axis

```{r, eval = FALSE}
ggviolin(
  data = bfam,
  y = "tot_cholesterol",
  x = "family",
  fill = "family",
  add = "mean_sd"
)
```

### Get n, mean, median, sd (1 pt)

```{r, eval = FALSE}
bfam |>
  group_by(family) |>
  get_summary_stats(tot_cholesterol,
    type = "common",
    show = c("mean", "median", "sd")
  )
```

### Is it normally distribute? (1 pt)

```{r, eval = FALSE}
ggqqplot(
  data = bfam,
  x = "tot_cholesterol",
  color = "family"
)


bfam |>
  group_by(family) |>
  shapiro_test(tot_cholesterol) |>
  gt()
```

> yes

### Is it variance similar between groups? (1 pt)

```{r, eval = FALSE}
b |>
  levene_test(tot_cholesterol ~ family) |>
  gt()
```

> yes

### What kind of test are you picking and why? (1 pt)

> anova_test since i think it is normally distribute and has equal variance

### 2. Declare null hypothesis $\mathcal{H}_0$

\mathcal{H}\_0\$ is that $family$ does not explain $tot\_cholesterol$ (1 pt)

### 3. Calculate test-statistic, exact p-value and plot (2 pt)

```{r, eval = FALSE}
bfam |>
  anova_test(tot_cholesterol ~ family) |>
  gt()


ggviolin(
  data = bfam,
  y = "tot_cholesterol",
  x = "family",
  fill = "family",
  add = "mean_sd"
) +
  stat_anova_test()
```

> My interpretation of the result
