---
title: "R Bootcamp Problem Set 4"
author: "Your name here"
date: last-modified 
---

## Problem Set

Use the data files in the `data/` directory to answer the questions.

For this problem set, you are allowed to help each other, but you are not
allowed to post correct answers in slack.

**The problem set is due 12pm on Sept 1.**

## Grading rubric

- Everything is good: full points
- Partially correct answer: depends on how many steps are correct
- Reasonable attempt: half points

## Question 1 **5 points**

1. Load the tidyverse and here packages.  
2. Import datasets: `data/data_rna_protein.csv.gz`.

`data_rna_protein.csv.gz`: This is a combined dataset from an RNAseq and SILAC
proteomics experiment, where a transcription factor (TF) was differentially
expressed and the fold change in RNA and protein calculated between
TF-expressing and non-expressing cells.
   
```{r, q1}
library(tidyverse)
library(here)

exp_tbl <- read_csv(
   here("problem-sets/data/data_rna_protein.csv.gz")
)
```

## Question 2 **5 points**

Using the imported data set, carry out the following:  

1. Inspect the data so you know what you are dealing with (`summary()` etc). 

2. Select only the following columns: `geneid`, `iDUX4_logFC`, `iDUX4_fdr`,
   `hl.ratio`, and `pval`.

3. Rename them as follows: rna_FC = iDUX4_logFC, rna_pval = iDUX4_fdr,
   protein_FC = hl.ratio, protein_pval = pval (hint: use `dplyr::rename()`)

4. Drop all rows with `NA` values in them (hint: use a function from `tidyr`)

6. Remove duplicate rows (hint: use `dplyr::distinct()`).

7. Arrange the table by descending `rna_FC` and ascending 
   `protein_FC`.
   
8. Conduct steps 2-7 by piping the output of one step to another (i.e, a single
   workflow & remember to comment).
   
9. Save the output of this workflow into a new object.

```{r, q2}
exp_tbl_subset <- exp_tbl |>
   select(geneid, iDUX4_logFC, iDUX4_fdr, hl.ratio, pval) |>
   rename(
      rna_FC = iDUX4_logFC,
      rna_pval = iDUX4_fdr,
      protein_FC = hl.ratio,
      protein_pval = pval
      ) |>
   drop_na() |>
   distinct() |>
   arrange(desc(rna_FC), protein_FC)
```

## Question 3  **5 points**

How well do the overall `rna_FC` and `protein_FC` values correlate in this
experiment?

Using the output from the above question, do the following:  

1. Create a scatter plot of `rna_FC` vs `protein_FC`. observe how the points scatter.

2. Add a line to the plot that would indicate perfect 1:1 correlation.
   Hint: use `geom_abline()` with its `slope` argument.
   
3. Add a linear model fit using `geom_smooth()` (`method = 'lm'`).
   Observe how the x=y line deviates from your geom_smooth line.

4. Calculate the Spearman correlation coefficient. (Hint: This uses a base R math
   function called `cor` - Use help() or Google to learn more and how to specify
   method as `spearman`)

5. Using all of the information from above, comment on the correlation between
   `rna_FC` and `protein_FC` below.

```{r, q4}
ggplot(exp_tbl_subset, aes(rna_FC, protein_FC)) +
   geom_point() +
   geom_abline(linewidth = 1, color = 'green') +
   geom_smooth(method = 'lm', se = FALSE, linewidth = 1)

rna_prot_cor <- cor(
   exp_tbl_subset$rna_FC,
   exp_tbl_subset$protein_FC,
   method = 'spearman'
)
```

### Answer

The green line indicates a perfect correlation, and the blue line is the linear model fit of the data. The Spearman correlation is `r round(rna_prot_cor, 3)`, indicating a strong positive correlation. One way to think about this is that there are `r round(rna_prot_cor, 3)` proteins made per mRNA.

## Submit

Be sure to click the "Render" button to render the HTML output.

Then paste the URL of this Posit Cloud project into the problem set on Canvas.
