---
title: "R Bootcamp Problem Set 2"
author: "Your Name Here"
date: last-modified
---

## Setup

Start by loading libraries you need analysis in the code chunk below.
When in doubt, start by loading the tidyverse package.

```{r, setup}
```

## Problem Set

Each problem below is worth **4 points**.

Use the data files in the `data/` directory to answer the questions.

For this problem set, you are allowed to help each other, but you are
not allowed to post correct answers in slack.

**The problem set is due 12pm on Aug 30.**

### Grading rubric

- Everything is good: 5 points
- Partially correct answers: 3-4 points
- Reasonable attempt: 2 points

## Question 1 

Import the data set `data_transcript_exp_subset` using the *readr*
package, and assign it to a variable called `exp_tbl`.

The file is located at the following path `data/data_transcript_exp_subset.csv.gz`

```{r, q1}
```

## Question 2

This data frame is a subset (100 lines) of transcript-level gene
expression data where transcript abundance was measured at two different
time points of a certain treatment conducted in triplicates.

The column names have the format of `molecule_time_replicate`

First, explore the structure of the data set using some of the functions
we learned in class (e.g., `summary()`)

```{r, q2}
```

Comment on whether this data set is tidy, and if not, list the
reasons why. Remember: in tidy data, every column represents a single
variable and every row represents a single observation

### Answer below

## Question 3 

**How will you reshape the data frame so that each row has only one experimental observation?**

Hint: Use `pivot_longer()`

```{r, q3}
```

## Question 4

**How will you modify the dataframe so that multiple variables are not present in a single column?**

Hint: Use `separate()`

```{r, q4}
```

## Question 5 

**How will you save your output as a TSV file?**

Hint: Use the *readr* cheatsheet to figure this out.

https://rstudio.cloud/learn/cheat-sheets

```{r, q5}
```
