---
title: "RNA Block - Problem Set 24"
---

## Problem Set

Total points: 20. Q1 - 10 pts, Q2,3 - 5 points each.

1. Perform differential expression analysis comparing DIV28 vs DIV0 (10 pts)

2. Are axonogenesis and cell cycle genes significantly differentially expressed? If so, in what direction (up/down-regulated)?  (5 pts)

3. Perform GSEA analysis using the Hallmark gene set. Make enrichment plot of `HALLMARK_G2M_CHECKPOINT` geneset.  (5 pts)


## Load libraries and generate gene information files (0 pts)

Make sure to run the code chunk below to load required libraries and generate `t2g` (file for `tximport`) and gene id to symbol mapping file.


```{r}
#| label: load-libs

library(biomaRt)
library(tximport)
library(tidyverse)
library(DESeq2)
library(ggrepel)
library(here)
library(cowplot)
library(rstatix)
library(clusterProfiler)
library(enrichplot)
library(msigdbr)
# run the code below to generate t2g file and gene id-symbol mapping file
mart <- biomaRt::useMart(
  "ENSEMBL_MART_ENSEMBL",
  dataset = "mmusculus_gene_ensembl"
)

t2g <- biomaRt::getBM(
  attributes = c(
    "ensembl_transcript_id",
    "ensembl_gene_id",
    "external_gene_name"
  ),
  mart = mart
) |>
  as_tibble()

# maps systematic to common gene names
gene_name_map <- t2g |>
  dplyr::select(-ensembl_transcript_id) |>
  unique()

```


## Q1 Perform differential expression analysis comparing DIV28 vs DIV0

### Prepare `metadata` and use `tximport` {.smaller}

```{r}
#| echo: true
#| eval: false


metadata <- data.frame(
  sample_id = list.files(here("??"),
                         pattern = "^DIV"),
  salmon_dirs = list.files(here("??"),
                           recursive = ??,
                           pattern = "quant.sf",
                           full.names = ??)
  ) |>
  separate(col = sample_id, into = c("timepoint","rep"), sep = "\\.", remove = F)

metadata$rep <- gsub(pattern = "Rep", replacement = "", metadata$rep) 

rownames(metadata) <- metadata$sample_id

metadata <- metadata |> 
  filter(timepoint %in% c("??","??")) 

salmdir <- metadata$salmon_dirs
names(salmdir) <- metadata$sample_id

txi <- tximport(
  files = salmdir,
  type = "salmon",
  tx2gene = t2g,
  dropInfReps = TRUE,
  countsFromAbundance = "lengthScaledTPM"
)
```

### Filter genes and perform `DESeq2` {.smaller}

```{r}
#| echo: true
#| eval: false


# examine distribution of TPMs
hist(log2(1 + rowSums(txi$??)), breaks = 40)

# decide a cutoff
keepG <- txi$abundance[log2(1 + rowSums(txi$abundance)) > ??,] |>
  rownames()

ddsTxi <- DESeqDataSetFromTximport(
  ??,
  colData = metadata,
  design = ~??
)

# keep genes with sufficient expession
ddsTxi <- ??

# run DESeq2
dds <- ??

# create a dataframe containing results and join w/gene symbols
diff <-
  results(
    dds,
    contrast = c("timepoint", "??", "??")
  ) |>
  # Change this into a dataframe
  as.data.frame() |>
  # Move ensembl gene IDs into their own column
  rownames_to_column(var = "ensembl_gene_id") |>
  # drop unused columns
  dplyr::select(-c(baseMean, lfcSE, stat, pvalue)) |>
  # Merge this with a table relating ensembl_gene_id with gene short names
  inner_join(gene_name_map) |>
  # Rename external_gene_name column
  dplyr::rename(gene = external_gene_name) |>
  as_tibble()
```

## Q2. Are axonogenesis and cell cycle genes significantly differentially expressed? If so, in what direction (up/down-regulated)?

```{r}
#| echo: true
#| eval: false
#| label: pathway-defined
#| fig.alt: "Description of the plot - PLEASE FILL IN"

cellcyclegenes <- ??(
  attributes = c("ensembl_gene_id"),
  filters = c("go_parent_term"),
  values = c("GO:0045787"),
  mart = mart
)


axongenes <- ??(
  attributes = c("ensembl_gene_id"),
  filters = c("go_parent_term"),
  values = c("GO:0007409"),
  mart = mart
)

diff_paths <-
  diff |>
  mutate(
    annot = case_when(
      ensembl_gene_id %in% axongenes$ensembl_gene_id ~ "??",
      ensembl_gene_id %in% cellcyclegenes$ensembl_gene_id ~ "??",
      .default = "none"
    )
  ) |>
  drop_na() # drop na

# Reorder these for plotting purposes
diff_paths$annot <-
  factor(
    diff_paths$annot,
    levels = c("none",
               "axonogenesis",
               "cellcycle")
  )

# calculate and report p-value using wilcox test
pvals <- wilcox_test(data = diff_paths,
                     ?? ~ ??,
                     ref.group = "??")

# make a plot of the LFC (y-axis) ~ pathways (x-axis)
ggplot(
  diff_paths,
  aes(
    x = ??,
    y = ??,
    fill = ??
  )
) +
  labs(
    x = "Gene class",
    y = "DIV28/DIV0, log2"
  ) +
  geom_hline(
    yintercept = 0,
    color = "gray",
    linetype = "dashed"
  ) +
  geom_boxplot(
    notch = TRUE,
    outlier.shape = NA
  ) +
  ylim(-5,5) +
  theme_cowplot() 
```



## Q3. Perform GSEA analysis using the Hallmark gene set.

```{r}
#| echo: true
#| eval: false
#| label: gsea


# retrieve mouse hallmark gene set from msigdb
mouse_hallmark <- msigdbr(species = "??") %>%
  filter(gs_cat == "??") %>% 
  dplyr::select(gs_name, gene_symbol)


# create a list of gene LFCs
rankedgenes <- diff %>% pull(??)

# add symbols as names of the list
names(rankedgenes) <- diff$??

# sort by LFC
rankedgenes <- sort(??, decreasing = TRUE)

# deduplicate
rankedgenes <- rankedgenes[!duplicated(names(rankedgenes))]


# run gsea
?? <- GSEA(geneList = ??,
                      eps = 0,
                     pAdjustMethod = "fdr",
                     pvalueCutoff = .05,
                     minGSSize = 20,
                     maxGSSize = 1000,
                     TERM2GENE = ??)


# plot "HALLMARK_G2M_CHECKPOINT"
gseaplot(x = ??, geneSetID = "??")
```

