---
title: "Problem Set Stats Bootcamp - class 12"
subtitle: "Hypothesis Testing"
author: "Neelanjan Mukherjee"
editor: visual
---

```{r}
#| echo: false
library(tidyverse)
library(rstatix)
library(ggpubr)
library(gt)
library(janitor)
library(here)
```

```{r}
biochem <- read_tsv("http://mtweb.cs.ucl.ac.uk/HSMICE/PHENOTYPES/Biochemistry.txt", show_col_types = FALSE) |>
  janitor::clean_names()

# simplify names a bit more
colnames(biochem) <- gsub(pattern = "biochem_", replacement = "", colnames(biochem))

# we are going to simplify this a bit and only keep some columns
keep <- colnames(biochem)[c(1, 6, 9, 14, 15, 24:28)]
biochem <- biochem[, keep]

# get weights for each individual mouse
# careful: did not come with column names
weight <- read_tsv("http://mtweb.cs.ucl.ac.uk/HSMICE/PHENOTYPES/weight", col_names = F, show_col_types = FALSE)

# add column names
colnames(weight) <- c("subject_name", "weight")

# add weight to biochem table and get rid of NAs
# rename gender to sex
b <- inner_join(biochem, weight, by = "subject_name") |>
  na.omit() |>
  rename(sex = gender)
```

## Problem \# 1

Is there an association between mouse calcium and sodium levels?

### 1. Make a scatterplot to inspect variable

```{r}
```

### 2. Are they normal (enough)?

```{r}
```

Which test will you use and why?

> 

### 3. Declare null hypothesis $\mathcal{H}_0$

$\mathcal{H}_0$ is that there is no dependency/association between $calcium$ and  $sodium$

### 4. Calculate and plot the correlation on a scatterplot

```{r}
```




## Problem \# 2

## Do mouse calcium levels explain mouse sodium levels? If so, to what extent?   

Use a linear model to do the following:  

### 1. Specify the Response and Explanatory variables --- (2 pts)  

> The response variable y is 
> The explantory variable x is 


### 2. Declare the null hypothesis --- (1 pts)

> The null hypothesis is ...

### 3. Use the `lm` function to create a fit (linear model) 


also save the slope and intercept for later

```{r}
#| label: lm fit
```

### 4. Add residuals to the data and create a plot visualizing the residuals

```{r}
#| label: augment and viz
```

### 5. Calculate the $R^2$ and compare to $R^2$ from fit

$R^2 = 1 - \displaystyle \frac {SS_{fit}}{SS_{null}}$

$SS_{fit} = \sum_{i=1}^{n} (data - line)^2 = \sum_{i=1}^{n} (y_{i} - (\beta_0 \cdot 1+ \beta_1 \cdot x)^2$

$SS_{null}$ — sum of squared errors around the mean of $y$

$SS_{null} = \sum_{i=1}^{n} (data - mean)^2 = \sum_{i=1}^{n} (y_{i} - \overline{y})^2$


  

```{r}
#| label: rsq
```


### 6. Using $R^2$, describe the extent to which calcium explains sodium levels

> 

### 7. Report (do not calculate) the $p-value$ and your decision on the null

```{r }
```

> The null hypothesis is ...

Calcium levels to predict sodium levels.

## Problem \# 3

What is the association between mouse weight and age levels for different sexes?


### 1. Calculate the pearson correlation coefficient between weight and age for females and males

```{r}
```

### 2. Describe your observations

> The relationship between weight and age...
