---
title: "R Bootcamp Problem Set 6"
author: "Your name here"
date: last-modified 
---

## Problem Set

Use the data files in the `data/` directory to answer the questions.

For this problem set, you are allowed to help each other, but you are not
allowed to post correct answers in slack.

**The problem set is due 12pm on Sept 1.**

## Grading rubric

- Everything is good: full points
- Partially correct answer: depends on how many steps are correct
- Reasonable attempt: half points

## Libraries

Load the libraries you need for analysis below.

```{r}
```

## Question 1 - **5 points**

Run the following chunk:

```{r}
#| label: run_me
set.seed(42)
x <- sample(1000, replace = TRUE)
```

Now use logical indexing to find the number of values > 450 in `x`.

```{r}
```

## Question 2 - **5 points**

Count the number of species in the `penguins` tibble using `forcats::fct_count()` 

Count number of island + sex combinations using `dplyr::count()`, and sort the result by count.

```{r}
```

```{r}
```

## Question 3 - **5 points**

Use `stringr::str_c()` to combine upper and lowercase letters from `letters` and `LETTERS` with a slash.

Your answer should look like: `"A/a" "B/b" "C/c"` etc.

```{r}
```

Use `stringr::str_split()` or one of its variants to split up the strings you made above and extract the letter after the slash.

```{r}
```

## Question 4 - **5 points**

Create a ggplot using the diamonds data set.

- Generate a `ggridges::geom_density_ridges()` for the prices, with a different `fill` color for each `cut`.
- Recolor the densities using `ggplot2::scale_fill_brewer()`, choosing a specific palette.
- make the outline of the densities black, and change their alpha to 0.2.
- change the theme to `cowplot::theme_minimal_grid()`
- remove the legend (google: "remove legend from ggplot2")
- add an informative title and subtitle using `ggplot2::labs()`.

```{r}
```

## Submit

Be sure to click the "Render" button to render the HTML output.

Then paste the URL of this Posit Cloud project into the problem set on Canvas.
