---
title: "DNA Block - Problem Set 21"
execute:
  eval: false
---

## Problem Set

In this problem set you'll examine the binding sites of the Reb1 transcription factor by CUT&RUN.

The first several chunks just run, putting libraries and files in your environment

Each question below is worth 4 points.

### Setup

Start by loading libraries you need analysis in the code chunk below.

```{r}
#| label: load-libs
#| message: false
library(tidyverse)
library(here)
library(valr)
library(cowplot)

# genome viz
library(TxDb.Scerevisiae.UCSC.sacCer3.sgdGene)
library(Gviz)
library(cpp11bigwig)

# motif discovery and viz
library(BSgenome.Scerevisiae.UCSC.sacCer3)
library(rGADEM)
library(seqLogo)
```

Next, load the coverage tracks for CUT&RUN data.

```{r}
#| label: make-tracks
track_start <- 1e5
track_end <- 2e5

# genes track
sgd_genes_trk <-
  GeneRegionTrack(
    TxDb.Scerevisiae.UCSC.sacCer3.sgdGene,
    chromosome = "chrII",
    start = track_start,
    end = track_end,
    fill = "#009E73",
    background.title = "white",
    col.title = "black",
    fontsize = 14
  )

# signal tracks
track_info <-
  tibble(
    file_name = c(
      "CutRun_Reb1_lt120.bw",
      "CutRun_Abf1_lt120.bw",
      "CutRun_Reb1_gt150.bw",
      "CutRun_Abf1_gt150.bw"
    ),
    sample_type = c(
      "Reb1_Short", "Abf1_Short",
      "Reb1_Long", "Abf1_Long"
    )
  ) |>
  mutate(
    file_path = here("data/block-dna", file_name),
    big_wig = purrr::map(
      file_path, ~ read_bigwig(.x, as = "gr")
    ),
    data_track = purrr::map2(
      big_wig, sample_type,
      ~ DataTrack(
          .x,
          name = .y,
          background.title = "white",
          col.title = "black",
          col.axis = "black",
          fontsize = 12
      )
    )
  ) |>
  dplyr::select(sample_type, big_wig, data_track)

# x-axis track
x_axis_trk <- GenomeAxisTrack(
  col = "black",
  col.axis = "black",
  fontsize = 16
)
```

Now that we have tracks loaded, we can make a plot.

```{r}
#| label: plot-tracks
plotTracks(
  c(
    sgd_genes_trk,
    track_info$data_track,
    x_axis_trk
  ),
  from = track_start,
  to = track_end,
  chromosome = "chrII",
  transcriptAnnotation = "gene",
  shape = "arrow",
  type = "histogram"
)
```

### Question 1

How do you interpret the differences in signals between the short and long fragments above? I.e.,
where are the short fragments enriched? And where do you see more of the long fragments?

> Answer.

### Question 2

Remake the plot above, but zoom in to a promoter region that has strong enrichment
for both Reb1 and Abf1 (short fragments).

Do the signals for Reb1 and Abf1 line up with one another? Why is this the case?

> Answer.

```{r}
#| label: plot-tracks-2
#| eval: false
plotTracks(
  c(
    sgd_genes_trk,
    track_info$data_track,
    x_axis_trk
  ),
  from = ___,
  to = ___,
  chromosome = "chrII",
  transcriptAnnotation = "gene",
  shape = "arrow",
  type = "histogram"
)
```

### Question 3

Next we'll take a look at Reb1 CUT&RUN data. In the following chunk, use the appraoch
we took in class to identify enriched sites of Reb1 binding.

```{r}
#| eval: false
#| label: peak-calling
reb1_tbl <- read_bigwig(here("data/block-dna/CutRun_Reb1_lt120.bw"))

# number of reads in the original Reb1 BED file
total_reads <- 16e6

genome <- read_genome(here("data/block-dna/sacCer3.chrom.sizes"))

# how can we calculate genome size?
genome_size <- ___

# define the genome-wide lambda value here
genome_lambda <- ___ / ___

peak_calls <-
  reb1_tbl |>
  # define single-base sites
  mutate(
    midpoint = start + round((end - start) / 2),
    start = midpoint,
    end = start + 1,
    # use the poisson to calculate a p-value with the genome-wide lambda
    pval = ___(score, genome_lambda),
    # convert p-values to FDR
    fdr = p.adjust(pval, method = "fdr")
  )
```

Let's take a look at a plot of the p-value across a chromosome.

Use `geom_hline()` to draw a red horizontal line at a cutoff that selects ~10 regions
enriched for Reb1 binding. You'll use this cutoff in the code chunks below.

```{r}
#| label: plot-peak-sig
#| eval: false
ggplot(
  filter(peak_calls, chrom == "chrII"),
  # convert p-values to positive values for plotting
  aes(start, ___)
) +
  geom_line() +
  xlim(track_start, track_end) +
  theme_cowplot() +
  geom_hline(___)
```

### Question 4

How many peaks are called in this region? Use the cutoff you defined above to identify
"peaks" of Reb1 binding.

> Answer.

How many *total* peaks are identified in the genome using this cutoff?

> Answer.

```{r}
#| label: n-peaks
#| eval: false
# most stringent cut-off
peak_calls_sig <-
  filter(
    peak_calls,
    ___
    ) |>
    # collapse neighboring, significant sites
    bed_merge(max_dist = 20)

filter(
  peak_calls_sig,
  chrom == "chrII" &
    start >= track_start &
    end <= track_end
)
```

Let's visualize these peaks in the context of genomic CUT&RUN signal. We need to define an `AnnotationTrack` with the peak intervals, which we can plot against the CUT&RUN coverage we defined above.

```{r}
#| label: plot-peaks
#| eval: false
# need a GRanges object to convert to an AnnotationTrack
peak_calls_gr <-
  GRanges(
    seqnames = peak_calls_sig$chrom,
    ranges = IRanges(peak_calls_sig$start, peak_calls_sig$end)
  )

peak_calls_trk <-
  AnnotationTrack(
    peak_calls_gr,
    name = "Peak calls",
    fill = "red",
    background.title = "white",
    col.title = "red",
    fontsize = 16,
    rotation.title = 0
  )

reb1_short_trk <-
  filter(
    track_info,
    sample_type == "Reb1_Short"
  ) |>
  pull(data_track)

plotTracks(
  c(
    sgd_genes_trk,
    reb1_short_trk,
    peak_calls_trk,
    x_axis_trk
  ),
  from = track_start,
  to = track_end,
  chromosome = "chrII",
  transcriptAnnotation = "gene",
  shape = "arrow",
  type = "histogram"
)
```

### Question 5

Use the peak calls you defined to identify a putative sequence motif bound by Reb1. You can
assume that the most abundant motif identified is the most likely candidate.

Use Google and Pubmed to identify a study that defines a Reb1 motif using a genomewide analysis. How well does your
motify match the previously defined one?

```{r}
#| label: motif-discovery
#| eval: false
peak_seqs <- BSgenome::getSeq(
  # provided by BSgenome.Scerevisiae.UCSC.sacCer3
  Scerevisiae,
  peak_calls_gr
)

# takes ~2 minutes to run
gadem <- rGADEM::GADEM(
  peak_seqs,
  genome = Scerevisiae,
  verbose = 1
)

# look at the consensus motifs
consensus(gadem)

# how many consensus motifs are there?
nOccurrences(gadem)
```

Now let's look at the sequence logo for the top hit.

```{r}
#| label: plot-logo
#| eval: false
pwm <- gadem@motifList[[1]]@pwm

seqLogo::seqLogo(pwm)
```
