---
title: "Exercises 7: Gene Expression Analysis with Tidyverse"
subtitle: "Advanced Modeling and Functional Analysis"
author: "Your Name"
date: today
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

In this problem set, you'll work with the Brauer gene expression dataset to practice comprehensive tidyverse skills including data tidying, transformation, joins, pivoting, string manipulation, and **statistical modeling using broom**. The dataset contains gene expression measurements for yeast genes under different nutrient limitations and growth rates.

## Predictions

Before we start tidying and analyzing the data, take a moment to predict what you might find.

- Question: What patterns do you expect to see in gene expression across different nutrients and growth rates?
- Hypothesis: Genes involved in nutrient uptake and metabolism will show higher expression under their respective limiting conditions.
- Hypohtesis: The different nutrient conditions will cause distinct gene expression profiles.

# Setup and Data Loading

## Load Required Libraries

```{r}
#| label: setup-01
#| message: false

# Load required libraries

```

## Load the Data

**Task 1:** Load the raw Brauer gene expression data and examine its structure. What makes this data "messy" or untidy?

*Breadcrumbs: Use `read_tsv()` to load the data from the URL. Examine column names and the first few rows. Think about tidy data principles - what issues do you see with the current format?*

```{r}
#|label: data-loading-02

# Load the Brauer gene expression data
url <- "https://github.com/rnabioco/molb-7950/raw/refs/heads/main/data/bootcamp/brauer_gene_exp_raw.tsv.gz"
```

```{r}
#| label: data-exploration-03

# Examine the structure of the data

```

# Part 1: Data Tidying (tidyr)

We want to create a tibble that looks like this:

```r
# A tibble: 199,296 × 4
   systematic_name nutrient  rate exp_level
   <chr>           <fct>    <dbl>     <dbl>
 1 YPR204W         Glucose   0.05      1.17
 2 YPR204W         Glucose   0.1       1
 3 YPR204W         Glucose   0.15      0.86
 4 YPR204W         Glucose   0.2       0.77
 5 YPR204W         Glucose   0.25      0.53
 6 YPR204W         Glucose   0.3       0.3
 7 YPR204W         Ammonia   0.05      2.79
 8 YPR204W         Ammonia   0.1       2
 9 YPR204W         Ammonia   0.15      0.6
10 YPR204W         Ammonia   0.2       0.16
# ℹ 199,286 more rows
# ℹ Use `print(n = ...)` to see more rows
```

*Note the classes of each column - systematic_name is character, nutrient is a factor, rate is numeric, and exp_level is numeric.*

## Create a Tidy Dataset

**Task 3:** Transform the wide-format expression data into a long format suitable for analysis.

*Breadcrumbs:*

- First select the relevant columns (`systematic_name` and the expression columns (`G0.05`, etc))
- Then use `pivot_longer()` to convert expression columns to rows
- The column names contain both nutrient type (`G`) and growth rate (`0.05`) information - use `separate_wider_position()` to split the first character (nutrient abbreviation) from the numeric rate. The key here is to specify `widths = c(nutrient_abbr = 1, rate = 4)` to split after the first character, and put the rest into `rate`. Not all the rates have 4 characters, so use `too_few = "align_start"` to handle that.
- Create a nutrient lookup table to convert abbreviations to full names. The lookup table should look like this:

  ```r
  # A tibble: 6 × 2
    nutrient_abbr nutrient
    <chr>         <chr>
  1 G             Glucose
  2 N             Ammonia
  3 P             Phosphate
  4 S             Sulfate
  5 L             Leucine
  6 U             Uracil
  ```

- Use that table with a `join` function to add full nutrient names to the tidied data.
- Convert the `rate` column to numeric and select the final columns in the desired order
- Remove any NA values from `systematic_name` and `exp_level` using `filter()`

```{r}
#| label: pivot-longer-05

```


# Part 2: Expression patterns

This is a "cheat" chunk to load the tidy data directly if you want to skip ahead. Otherwise, use the code above to create the tidy data from the raw data.

```{r}
#| label: tidy-data-skip-ahead
#| echo: true
#| eval: false
brauer_tidy_tbl <- read_tsv(
  "https://github.com/rnabioco/molb-7950/raw/refs/heads/main/data/bootcamp/brauer_gene_exp_tidy.tsv.gz"
)

yeast_gene_info_tbl <- read_tsv(
  "https://github.com/rnabioco/molb-7950/raw/refs/heads/main/data/bootcamp/yeast_go_terms.tsv.gz"
)

yeast_protein_props_tbl <- read_tsv(
  "https://github.com/rnabioco/molb-7950/raw/refs/heads/main/data/bootcamp/yeast_prot_prop.tsv.gz"
)
```

## Explore Expression Patterns

**Task 6:** Calculate summary statistics for gene expression by nutrient type.

*Breadcrumbs:*
- Use `group_by()` and `summarize()` to calculate mean, median, and standard deviation of expression values for each nutrient. Which nutrients show the highest variability in expression?*

```{r}
#| label: expression-summary-09
# Calculate summary statistics by nutrient

```

```{r}
#| label: extreme-expression-10
#| message: false
# Find genes with extreme expression values

```

Inspect the results and note any patterns you observe in high and low expression genes across different nutrient-rate combinations.

Next, make a boxplot to visualize the distribution of expression levels for each nutrient condition. What insights can you draw from the plot?

```{r}
#| label: extreme-expression-10-boxplot
#|
```

## Identify High and Low Expression

**Task 7:** Find genes with extreme expression values under different conditions.

*Breadcrumbs: For each nutrient-rate combination, identify the top 5 highest and lowest expressing genes. Use `slice_max()` and `slice_min()` or ranking functions. What patterns do you notice?*

```{r}
#| label: extreme-expression-11
# Find genes with extreme expression values

```


