---
title: "R Bootcamp Problem Set 4"
author: "Your name here"
date: last-modified
execute:
  eval: false
format:
  html:
    embed-resources: true
---

## Problem Set

Use the data files in the `data/` directory to answer the questions.

For this problem set, you are allowed to help each other, but you are not
allowed to post correct answers in slack.

**The problem set is due 5pm on Aug 29**

## Grading rubric

- Everything is good: full points
- Partially correct answer: depends on how many steps are correct
- Reasonable attempt: half points

## Question 1 **5 points**

1. Load the tidyverse and here packages using `library()`.
2. Import datasets: `data/data_rna_protein.csv.gz` using `read_csv()` and `here()`.

`data_rna_protein.csv.gz`: This is a combined dataset from an RNAseq and SILAC
proteomics experiment, where a transcription factor (TF) was differentially
expressed and the fold change in RNA and protein calculated between
TF-expressing and non-expressing cells.

```{r}
#| label: q1
#| echo: true
library(tidyverse)
library(here)

exp_tbl <- read_csv(
  here(___)
)
```

## Question 2 **5 points**

Let's build a data processing workflow step by step. This teaches you how to build complex pipelines gradually - a key skill in data analysis.

**Step 1**: First, explore the data so you know what you're working with. Use `glimpse()` to see column types and `summary()` to see distributions:

```{r}
#| label: q2-explore
#| echo: true
# Always explore your data first!
```

**Step 2**: Select only the columns we need:

- `geneid`      (gene identifier)
- `iDUX4_logFC` (RNA fold change)
- `iDUX4_fdr`   (RNA pvalue)
- `hl.ratio`    (protein fold change)
- `pval`        (protein pvalue)

Use `select()` and list the columns you want to keep:

```{r}
#| label: q2-select
#| echo: true
exp_tbl |>
  select(___)
```

**Step 3**: Rename columns for clarity (this makes your code more readable).

Use `dplyr::rename()` with the pattern `new_name = old_name, ...`:

```{r}
#| label: q2-rename
#| echo: true
exp_tbl |>
  select(___) |>
  rename(
    ___ = ___,
    # etc
  )
```

**Step 4**: Clean the data by removing rows with missing values. Use `drop_na()` to remove rows with any missing values, and `distinct()` to remove duplicate rows:

```{r}
#| label: q2-clean
#| echo: true
exp_tbl |>
  select(___) |>
  rename(
    ___ = ___,
    # etc
  ) |>
  ___() |> # Remove rows with any missing values
  ___() # Remove duplicate rows
```

**Step 5**: Finally, arrange the data and save it. Use `arrange()` to sort by RNA fold change (high to low), then protein fold change (low to high):

```{r}
#| label: q2-final
#| echo: true
exp_tbl_subset <- exp_tbl |>
  select(___) |>
  rename(
    ___ = ___,
    # etc
  ) |>
  ___() |> # Remove rows with any missing values
  ___() |> # Remove duplicate rows
  # Sort by RNA fold change (high to low), then protein fold change (low to high)
  ___(___, ___)

exp_tbl_subset
```

## Question 3 **5 points**

Let's practice good data analysis habits by checking for potential issues. Quality control is essential in real data analysis.

**Check for duplicates and missing values**:

1. Use `count()` to check for duplicate genes
1. Use `summarize()` with `across()` to count missing values in all columns
1. Use summary statistics to understand data distributions

```{r}
#| label: q4-quality-check
#| echo: true
# Check for duplicate genes (there shouldn't be any after distinct())
exp_tbl_subset |>
  count(___) |>
  ___(n > 1) # Any genes appearing more than once?
```

```{r}
#| label: q4-quality-check-2
#| echo: true
# Summary of missing values by column
exp_tbl_subset |>
  summarize(
    # first blank select variables
    # second blank applies a function to count NA values
    across(___, ___)
  )
```

```{r}
#| label: q4-quality-check-3
#| echo: true
# Look at the distribution of our main variables
exp_tbl_subset |>
  summarize(
    across(
      # specify the variables to summarize
      ___,
      list(
        # mean
        mean = ~ mean(., na.rm = TRUE),
        # now do median
        ___ = ~ ___(., na.rm = TRUE),
        # and sd
        ___ = ~ ___(., na.rm = TRUE)
      )
    ),
    .groups = "drop"
  )
```

## Question 4 **5 points**

How well do the overall `rna_FC` and `protein_FC` values correlate in this experiment? We'll explore this with visualization and statistics.

**Step 1**: Create a scatter plot of `rna_FC` vs `protein_FC` using `ggplot()`. Use:

- `aes()` to map x and y variables
- `geom_point()` to create the scatter plot
- `labs()` to add informative axis labels and title

```{r}
#| label: q3-plot
#| echo: true
ggplot(
  ___,
  aes(
    x = ___,
    y = ___
  )
) +
  # ad points
  ___() +
  # add labels
  labs(
    x = "___",
    y = "___",
    title = "___"
  )
```

**Step 2**: Add reference lines to help interpret the correlation. Use:

- `geom_abline(slope = 1, intercept = 0)` for perfect correlation line
- `geom_smooth(method = "lm", se = FALSE)` for the computed trend line
- adjust the `geom_point()` aesthetic to `alpha = 0.6`, making points slightly transparent for better visualization

```{r}
#| label: q3-plot-lines
#| echo: true
ggplot(
  ___,
  aes(
    x = ___,
    y = ___
  )
) +
  # Add transparent points (change the ???)
  geom_???(alpha = 0.6) +
  # Add the perfect correlation line (change the ???)
  geom_???(slope = 1, intercept = 0, color = "red", linewidth = 1) +
  # Add the computed trend line (change the ???)
  geom_???(method = "lm", se = FALSE, color = "blue", linewidth = 1) +
  labs(
    x = "___",
    y = "___",
    title = "___"
  )
```

**Step 3**: Calculate the correlation coefficient using `cor()`. Use Spearman correlation since it's robust to outliers. Use `?cor` to see the function documentation. You will need to specify two vectors for the calculation, and it's easiest to provide them using the `$` operator to extract columns from the data frame.

```{r}
#| label: q3-correlation
#| echo: true
rna_prot_cor <- cor(
  # specify the first vector
  ___,
  # specify the second vector
  ___,
  method = "spearman"
)

rna_prot_cor
```

### Answer

**[ YOUR ANSWER HERE ]**

## Submit

Be sure to click the "Render" button to render the HTML output.

Then paste the URL of this Posit Cloud project into the problem set on Canvas.
