---
title: "R Bootcamp Problem Set 1"
---

## Problem Set

Each problem below is worth **10 points**.

**The problem set is due 12pm on Aug 30.**

### Grading rubric

- Everything is good: 10 points
- Partially correct answers: 6-8 points
- Reasonable attempt: 4 points

## Setup

Start by loading libraries you need analysis in the code chunk below.
When in doubt, start by loading the tidyverse package.

```{r}

```

## Question 1

Create 3 different vectors called `x`, `y`, and `z`:

- `x` should be character vector of length 5
- `y` should be a numerica vector of length 5
- `z` should be a logical vector of length 5

Use `length()` to calculate the length of each vector.

```{r}

```

## Question 2

Using the vectors you created above, create a new tibble with column names
`x`, `y`, and `z`.

Use `nrow()` and `ncol()` to calculate the number of rows and columns.


```{r}

```

What do you notice about the length of the vectors and the number of rows?

### Answer here

## Submit

Be sure to click the "Render" button to render the HTML output.

Then paste the URL of this Posit Cloud project into the problem set on Canvas.
