---
title: "Problem Set Stats Bootcamp - class 12"
subtitle: "Hypothesis Testing"
author: "Neelanjan Mukherjee"
editor: visual
---

```{r}
#| echo: false
library(tidyverse)
library(rstatix)
library(ggpubr)
library(gt)
library(janitor)
library(here)
```

```{r}
biochem <- read_tsv("http://mtweb.cs.ucl.ac.uk/HSMICE/PHENOTYPES/Biochemistry.txt", show_col_types = FALSE) |>
  janitor::clean_names()

# simplify names a bit more
colnames(biochem) <- gsub(pattern = "biochem_", replacement = "", colnames(biochem))

# we are going to simplify this a bit and only keep some columns
keep <- colnames(biochem)[c(1, 6, 9, 14, 15, 24:28)]
biochem <- biochem[, keep]

# get weights for each individual mouse
# careful: did not come with column names
weight <- read_tsv("http://mtweb.cs.ucl.ac.uk/HSMICE/PHENOTYPES/weight", col_names = F, show_col_types = FALSE)

# add column names
colnames(weight) <- c("subject_name", "weight")

# add weight to biochem table and get rid of NAs
# rename gender to sex
b <- inner_join(biochem, weight, by = "subject_name") |>
  na.omit() |>
  rename(sex = gender)
```

## Problem \# 1

Can mouse sex explain mouse cholesterol? {.smaller}

## STEP 1: Null hypothesis and variable specification

$\mathcal{H}_0:$

> ?? is the response variable

> ?? is the explanatory variable

## STEP 2: Fit linear model and examine results {.smaller}

```{r}
#| echo: true
```

Fit summary:

```{r}
#| echo: true
#| output-location: column-fragment
```

Coefficient summary:

```{r}
#| echo: true
#| output-location: column-fragment
```

## Collecting residuals and other information {.smaller}

add residuals and other information

```{r}
#| echo: true
#| output-location: column-fragment
```

## STEP 4: Visualize the error around fit {.smaller}

```{r}
#| echo: true
#| output-location: column-fragment


# plot of data with mean and colored by residuals
```

## STEP 3: Visualize the error around the null (mean weight) {.smaller}

```{r}
#| echo: true
#| output-location: column-fragment
```

## Plot the fit error and the null error as 2 panels {.smaller}

```{r}
#| echo: true
```

## Calculate $R^2$ {.smaller}

```{r}
#| echo: true
#| output-location: column-fragment
```

$R^2 = 1 - \displaystyle \frac {SS_{fit}}{SS_{null}}$

```{r}
#| echo: true
#| output-location: column-fragment
```

check agains Rsq in your fit

```{r}
#| echo: true
#| output-location: column-fragment
```

## Compare to traditional t-test {.smaller}

```{r}
#| echo: true
#| output-location: column-fragment
```

```{r}
#| echo: true
#| output-location: column-fragment
```

## Provide your interpreation of the result
