---
title: "Problem Set Stats Bootcamp - class 10"
subtitle: "Distributions and probabilities"
author: "Neelanjan Mukherjee"
editor: visual
---

```{r}
#| echo: false
library(tidyverse)
library(rstatix)
library(cowplot)
```

## Problem \# 1

Assume that the height [jackalopes](https://www.youtube.com/watch?v=VkUi4qdZStQ) fit a normal distribution. Through careful field research measuring 1000 wild jackalopes, we have determined the mean height is 97 cm and the standard deviation is 10 cm. Your was camping and found a jackalope. Being a great friend and knowing your interest in jackalopes, they (harmlessly) subdued and measured the wild jackalope and found that it was 75 cm.

1.  Simulate a normal distribution of `1000` jackalope heights using the `mean` and `sd` you painstakingly measured.

```{r}
```

2.  Plot the density of the jackalope height distribution. Indicate with a vertical line the height of the jackalope your friend measured.

```{r}
```

3.  Calculate the probability of a jackalope being 75 cm or shorter.

```{r}
```

3.  Are jackalope heights normally distributed?

```{r}
```

