---
title: "Problem Set Stats Bootcamp - class 12"
subtitle: "Hypothesis Testing"
author: "Neelanjan Mukherjee"
editor: visual
---

```{r}
#| echo: false
library(tidyverse)
library(rstatix)
library(ggpubr)
library(gt)
library(janitor)
library(here)
```

```{r}
#| label: prep biochem
biochem <- read_tsv("http://mtweb.cs.ucl.ac.uk/HSMICE/PHENOTYPES/Biochemistry.txt", show_col_types = FALSE) |>
  janitor::clean_names()

# simplify names a bit more
colnames(biochem) <- gsub(pattern = "biochem_", replacement = "", colnames(biochem))

# we are going to simplify this a bit and only keep some columns
keep <- colnames(biochem)[c(1, 6, 9, 14, 15, 24:28)]
biochem <- biochem[, keep]

# get weights for each individual mouse
# careful: did not come with column names
weight <- read_tsv("http://mtweb.cs.ucl.ac.uk/HSMICE/PHENOTYPES/weight", col_names = F, show_col_types = FALSE)

# add column names
colnames(weight) <- c("subject_name", "weight")

# add weight to biochem table and get rid of NAs
# rename gender to sex
b <- inner_join(biochem, weight, by = "subject_name") |>
  na.omit() |>
  rename(sex = gender)
```

## Problem \# 1

Does mouse sex explain mouse total cholesterol levels? Make sure to run chunks above.

### 1. Examine and specify the variable(s) (1 pt)

> The response variable y is $??$\
> The explantory variable x is $??$

### Make a violin plot: (2 pt)

response variable on the y-axis

explanatory variable on the x-axis

```{r}
```

### Get n, mean, median, sd (1 pt)

```{r}
```

### Is it normally distribute? (1 pt)

```{r}
```

> Answer here

### Is it variance similar between groups? (1 pt)

```{r}
```

> Answer here

### What kind of test are you picking and why? (1 pt)

> Answer here

### 2. Declare null hypothesis $\mathcal{H}_0$ (1 pt)

$\mathcal{H}_0$ is that $??$ does not explain $??$

### 3. Calculate test-statistic, exact p-value and plot (2 pt)

```{r}
```

> My interpretation of the result

```{r}
#| label: prep fam
# i have pre-selected some families to compare
myfams <- c(
  "B1.5:E1.4(4) B1.5:A1.4(5)",
  "F1.3:A1.2(3) F1.3:E2.2(3)",
  "A1.3:D1.2(3) A1.3:H1.2(3)",
  "D5.4:G2.3(4) D5.4:C4.3(4)"
)

# only keep the familys in myfams
bfam <- b |>
  filter(family %in% myfams) |>
  droplevels()

# simplify family names and make factor
bfam$family <- gsub(pattern = "\\..*", replacement = "", x = bfam$family) |>
  as.factor()


# make B1 the reference (most similar to overall mean)
bfam$family <- relevel(x = bfam$family, ref = "B1")
```

## Problem \# 2

Does mouse family explain mouse total cholesterol levels? Make sure to run chunk above.

### 1. Examine and specify the variable(s) (1 pt)

> The response variable y is $??$\
> The explanatory variable x is $??$

### Make a plot: (2 pt)

response variable on the y-axis

explanatory variable on the x-axis

```{r}
```

### Get n, mean, median, sd (1 pt)

```{r}
```

### Is it normally distribute? (1 pt)

```{r}
```

> Answer here

### Is it variance similar between groups? (1 pt)

```{r}
```

> Answer here

### What kind of test are you picking and why? (1 pt)

> Answer here \### 2. Declare null hypothesis $\mathcal{H}_0$

### $\mathcal{H}_0$ is that $??$ does not explain $??$ (1 pt)

### 3. Calculate test-statistic, exact p-value and plot (2 pt)

```{r}
```

> My interpretation of the result
